Vragenlijst thesis evaluatie

### 2015-06-21 Thesisfollowup/ThesisFollowUpSophieMarien.tex
hoofdstuk worm propagatie aangevuld

Thesis follow up vragen van Maart en April terug beantwoord. 
\section{Vergelijk wat je gedaan hebt met het opgestelde werkplan. Hoever sta je? Moet het werkplan aangepast worden?}
Ik moet een tandje bijsteken en al verder beginnen schrijven. Ik was vergeten in mijn werkplan bij te steken dat ik een week op Athens ging gaan. 
\section{Schrijf een eerste versie van een hoofstuk over een afgewerkte bijdrage van je masterproef.}
Dit is meegegeven in bijlage. Ik heb ineens heel mijn voorlopige thesistekst in bijlage gestoken. 
\section{Geef een korte inhoud van het tweede artikel dat je gaat schrijven. (Dit is niet van toepassing voor de studenten van de Master Toegepaste Informatica)}
Het tweede artikel gaat gaan over mijn implementatie van FlipIt op een bedrijfsnetwerk. Eerst bespreek ik hoe ik FlipIt aanpas. Daarna stel ik een methode voor via matrices om de propagatie van een virus te beschrijven. De methode die gebruikt wordt is gelijkaardig aan die van Google Page Ranking. Via deze functie kan ik de totale controle berekenen van een speler over een spel. In het artikel leg ik de methode uit aan de hand van een default spel. 
Ook geef ik een eerste setting van een spel en de bijhorende benefit formules waaruit ik al dan niet een equilibrium kan bepalen.
\section{Beschrijf in 10 lijnen de demo die je bij je verdediging zult geven.}
Ik heb een theoretische thesis dus kan ik moeilijk een demo geven. Ik zal enkele voorbeelden uitwerken om mijn werkingsmethode uit te leggen.
\section{Wat is er afgesproken met je begeleider en/of promotor om op te leveren als major milestone die het praktische werk van je thesis afsluit?}
Dit is nog niet afgesproken.
\section{Evalueer de tijd die je in je thesis gestoken hebt in de voorbije 4 weken: veel te weinig / te weinig / voldoende / ruim voldoende / meer dan verwacht.}
De voorbije weken heb ik meer dan verwacht aan mijn thesis gewerkt.  
\section{Evalueer de vorderingen van je thesis t.o.v. het bereiken van de vooropgestelde doelstelling: ver achter op het schema / achter op het schema / goed op schema / vooruit op het schema / ver vooruit op het schema.}
Ik loop een beetje achter op schema want ik had toch graag nu al een aantal formuleringen bepaald en een model uitgewerkt.
\section{Stuur de ingevulde timesheet mee als bijlage.}

>>>>

Thesis follow up vragen van Maart en April terug beantwoord. 
\section{Hoever sta je? Moet het werkplan aangepast worden?}
 Momenteel moet ik mijn tekst verfijnen van het hoofdstuk over worm propagatie en de related work. De nash equilibria moeten ook nog berekent worden. Na de examens ga ik terug verder werken aan het berekenen van de Nash equilibria. Ondertussen werk ik elke avond een deel van de tekst af. De hoofdstukken tot en met het Nash Equilibria zijn al volledig uitgeschreven. 
\section{Wat zijn de belangrijkste elementen uit het besluit dat je gaat schrijven in je tekst?}
In mijn besluit ga ik, als ze bestaan, kunnen vaststellen wat de nash equilibria zijn en hier conclusies kunnen uittrekken voor de verdedigings-strategie"en van de defender. 
\section{Verfijn de outline van je thesistekst. Geef per hoofdstuk de subsecties, en per subsectie het kernidee en een schatting van het aantal bladzijdes van die subsectie.}

\begin{enumerate}
\item \textbf{Introduction} 
\begin{itemize}
\item Research questions
\item Contributions and results
\item Conjunctures and open problems
\item Overview thesis
\end{itemize}
\item \textbf{General context}
\begin{itemize}
\item[2.1] What is security
\item[2.2] A brief introduction to game theory
\item[2.3] The FlipIt game
\item[2.4] Extensions on FlipIt
\end{itemize}
\item \textbf{FlipIt with propagation delay}
\begin{itemize}
\item[3.1] Introduction
\item[3.2] Difference between FlipIt with and without propagation delay
\begin{itemize}
\item Single resource
\item Only one action per player
\item Immediate effect of the move
\item Stealth character of the moves
\item Cost associated with the move
\item Strategies
\item Application: Wormlike ATP
\end{itemize}
\item[3.3] Playing periodically with propagation delay
\end{itemize}
\item \textbf{Nash Equilibria}
\begin{itemize}
\item[4.1] Determining piecewise functions $opt_{D}(\delta_{A}$
\item[4.2] Determining piecewise functions $opt_{A}(\delta_{D}$
\item[4.3] Determing the Nash equilibria
\end{itemize}
\item\textbf{ Worm propagation}
\begin{itemize}
\item[5.1] Methods of worm propagation
\item[5.2] Graph models
\item[5.3] Models for worm propagation
\item[5.4] Method for calculating propagation delay through matrices
\end{itemize}
\item \textbf{Related Work}
\item \textbf{Conclusion}
\end{enumerate}

\section{Beschrijf in 10 lijnen de demo die je bij je verdediging zult geven.}
Ik heb een theoretische thesis dus kan ik moeilijk een demo geven. Ik zal enkele voorbeelden uitwerken om mijn werkingsmethode uit te leggen.
\section{Wat is er afgesproken met je begeleider en/of promotor om op te leveren als major milestone die het praktische werk van je thesis afsluit?}
Dit is besproken met de begeleiders maar er is niets afgesproken omdat op voorhand niet kan gezegd worden of de analyse van het FlipIt spel voldoende zal zijn omdat de indruk kan ontstaan dat dit resultaat eenvoudig te verkrijgen was. Er is dus niet afgesproken wat ik moet inleveren om mijn thesis af te sluiten.
%Als ik hier niet mee akkoord dan moet ik dit ook verduidelijken, en moet ik mijn standpunt verdedigen (in mijn tekst, en op de verdediging). 
Voorlopig werk ik verder af wat ik zelf al voor ogen had; namelijk het vinden van de nash equilibria en het schrijven van een related work over worm propagatie. \\
Mijn promotor stelde het volgende: Mijn masterproef is in de eerste plaats mijn verantwoordelijkheid, ik ben verantwoordelijk voor wat ik indien en wat ik doe. De rol van mijn begeleiders en promotor is hierbij advies en richtlijnen geven.
Als ik het hiermee niet eens ben dan zorg ik voor de nodige argumenten en stel een sterke tekst op samen met een goede verdediging van mijn werk. 
\section{Evalueer de tijd die je in je thesis gestoken hebt in de voorbije 4 weken: veel te weinig / te weinig / voldoende / ruim voldoende / meer dan verwacht.}
De voorbije weken heb ik meer dan verwacht aan mijn thesis gewerkt. Momenteel ben ik al aan het aantal uren gekomen dat ik aan mijn thesis moet werken. Hierbij zullen waarschijnlijk nog eens 150 uren bijkomen. 
\section{Evalueer de vorderingen van je thesis t.o.v. het bereiken van de vooropgestelde doelstelling: ver achter op het schema / achter op het schema / goed op schema / vooruit op het schema / ver vooruit op het schema.}
Ik zit achter op schema want anders had ik al in eerste zit kunnen afgeven. De komende week probeer ik mijn Nash equilibria hoofdstuk af te werken en daarna focus ik mij op het schrijven van de hoofdstukken.
\section{Stuur de ingevulde timesheet mee als bijlage.}

### 2015-06-21 doc/template/chap-4.tex
hoofdstuk worm propagatie aangevuld

The spreading of malware has already been extensively researched. Because of all the different types of propagation it is hard to pick out one model that can model all of them. Modelling the spread of malware depends on two great factors: the method used for the propagation and the graph where the malware will spread itself. Viruses and worms are the types of malware that are the most researched. \\
In this chapter we focus on the propagation models of worms. The propagation of a worm is not depending on the human interaction as it is with viruses. In section [] we present an overview of the most used propagation methods by worms. Next, in section [] different kind of graphs to model a computer network are introduced. Section [] lists all the kind of models to determine the time a worm needs to infect a sufficient number of nodes. These results can be used for the completion of parameter \textit{d}. Finally we introduce an easy method to calculate the delay of the propagation of a worm. 
\subsection{Introduction to this chapter}

Advanced Persistent Threads propagation koppelen met die van een worm.
A virus infects one node on the network and has to wait for human interaction to spread. So the spreading speed depends on the human interaction. 
\subsection*{Methods of propagation}

\begin{description}
\item Random scanning: Some worms propagate by using the method of random scanning 
IP addresses.  Rate of terms of succeeded randwomly choosen IP adresses -> maximise good ip adresses. After a while the remaining nodes will be les likely to be reached. An example of such a worm is `Code Red' .
\item Routing worms: they use BGP routing tables to scan the routable address space. 3 times faster than traditional worm that uses random scanning. Spyb0t or network bluepill.
\item email worms: use the email systems to propagate. The Kak worm is a Javascript computer worm that spread itself by exploiting a bug in Outlook Express. 
%Pikachu worm: The virus was mainly spread through Microsoft Outlook email attachments. The email containing the attached virus propagated through infected users by sending itself to all contacts in the user's Outlook address book. [5]
\item self stopping worm: reduces speed to avoid detection. Atak worm or self stopping worm
\end{description}

The start of such a propagation can be different. Either it is a targeted attack or a random attack. For example a usb stick or other shared resources can be used to start the spreading. It can be given to some special person or it can just be left behind in a parking lot. This can still be a targeted attack if it is the parking lot of a targeted company but there is more insecurity of where the usb stick will end. Another way is by sending an email. It can also start from one computer that is connected directly to the internet. Downloading malicious programs.

\subsection*{Models for worm propagation}
To use the models for worm propagation in our model of the FlipIt game it is suffi"ent to know the propagation speed of the worm and define after how many time units the network is defected. This will be equal to the \textit{d} parameter in our model.  
Stackelberg game: first move is from attacker, defender is follower.

@@

What do we need for an algorithm
\begin{description}
\item Graph network $G = < V, E>$
\item Graph matrix $[A]$ which is $|V| \times |V| $
\item Attack vector $[X]$ which is $1 \times |V|$
\item cummulative matrix $[M]$ which is $|V| \times |V|$
\item state matrix $[T]$  which is $|V| \times |V|$
\item Reset vector $[R]$
\item duration \textit{d}
\item time \textit{n}
\item rate $\delta _{0}$ of defender and $\delta _{1}$ of attacker
\end{description}

Initialisation algorithm:

\begin{verbatim}
initialisatie
	d=0
	A=basismatrix
	M=A^{0}
	n=0
	\delta_{0}
	\delta_{1}
	X
	R
	controller = defender

	Algorithm
	n:= n + 1;
	Check who is in control? ( through modulo )
	if ( defender & controller=defender)
				d:= d + 1;

	if ( defender & controller=attacker )
				G = X \times R  (flippen ten voordele van defender)
				d = 0
				controller = defender

	if ( attacker & controller=defender )
				controller=attacker
				..

	if ( attacker & contoller=attacker )
				d:= d + 1
				M = M x A
				T = T + M
				G = X x T

\end{verbatim}

>>>>

The spreading of malware has already been extensively researched. Because of all the different types of propagation it is hard to pick out one model that can model all of them. Modelling the spread of malware depends on two great factors: the method used for the propagation and the graph where the malware will spread itself. Viruses and worms are the types of malware that are the most researched. \\
In this chapter we focus on the propagation models of worms. The propagation of a worm is not depending on the human interaction as it is with viruses. In section [] we present an overview of the most used propagation methods by worms. Next, in section [] different kind of graphs to model a computer network are introduced. Section [] lists all the kind of models to determine the time a worm needs to infect a sufficient number of nodes. These results can be used for the completion of parameter \textit{d}. Finally we introduce an easy method to calculate the delay of the propagation of a worm.

This related work is based on the following paper: [importance journal], 
Advanced Persistent Threads propagation koppelen met die van een worm.

\subsection{Methods of propagation}

There exist two kinds of APT's in the context of propagation. There are APT's that only launch an attack on just one target. The mechanism they use to propagate is the dropper mechanism. The dropper is the initial attack vector that compromises the system. The other kind of APT's target multiple victims. They also use a dropper mechanism and have an additional mechanism for self-propagation. It can be either the virus propagation or the worm like propagation. A virus infects one node on the network and has to wait for human interaction to spread. So the spreading speed depends on the human interaction. If an APT uses a worm propagation method it will spread by itself after it has been dropped on the network. Because it is not easy to incorporate the human factor in a spreading method we will only examine the worm propagation models for our APT. \\

Below is a list of several common propagation methods:

\begin{description}
\item Scanning methods: A worm can use distinct ways of scanning: random, localized, topological or hit-list scanning.
\begin{itemize}
\item Random scanning: Some worms propagate by using the method of random targeting IP addresses. Rate of terms of succeeded randomly chosen IP addresses is very low. An example of such a worm is `Code Red' ans 'Slammer'.
\item Localized scanning: --Localized scanning,
used by the Code Red II and Nimda worms, preferentially
scans for hosts in the ''local'' address space.--
\item Topological scanning: --Topological scanning, used by the Morris worm, relies
on the ''address'' information contained in the victim
machines to locate new targets.--
\item Hit-list scanning: With the method of hit-list scanning, before hand a short-list of vulnerable systems are made. --to speed up
the spread of worms at the initial stage. This list consists
of potentially vulnerable machines that are gathered
beforehand and targeted first when the worm
is released. An extreme case for the hit-list-scanning
worms is a flash worm, which gathers all vulnerable
machines into the list.--
\end{itemize}   
\item Routing worms: they use BGP routing tables to scan the routable address space. 3 times faster than traditional worm that uses random scanning. Spyb0t or network Bluepill. --a ''routing worm'' [20]. Zou et al. designed two types
of routing worms [20]. One type, based on Class-A
(x.0.0.0/8) address allocations, is thus called 'ClassA
routing worms.'' Such worms can reduce the scanning
space to 45.3\% of the entire IPv4 address space.
The other type, based on BGP routing tables, is thus
called''BGP routing worms'' Such worms can reduce
the scanning space to only about 28.6\% of the entire
IPv4 address space.--
\item email worms: use the email systems to propagate. The Kak worm is a Javascript computer worm that spread itself by exploiting a bug in Outlook Express.  GhostNet
\item DNS random scanning: --Another strategy that a worm
can potentially employ is DNS random scanning [5],
in which a worm uses the DNS infrastructure to locate
likely targets by guessing DNS names instead of
IP addresses. Such a worm in the IPv6 Internet is
shown to exhibit a propagation speed comparable to
that of an IPv4 random-scanning worm--
%Pikachu worm: The virus was mainly spread through Microsoft Outlook email attachments. The email containing the attached virus propagated through infected users by sending itself to all contacts in the user's Outlook address book. [5]
\item self stopping worm: reduces speed to avoid detection. Atak worm or self stopping worm
\item network sharing Shamoon
\item shared files or spreading through SMB (shared message block or  Common Internet File System (CIFS)) Flame
\item zero-day vulnerabilities
\end{description}

--dit stuk tekst overbodig--
The start of such a propagation can be different. Either it is a targeted attack or a random attack. For example a usb stick or other shared resources can be used to start the spreading. It can be given to some special person or it can just be left behind in a parking lot. This can still be a targeted attack if it is the parking lot of a targeted company but there is more insecurity of where the usb stick will end. Another way is by sending an email. It can also start from one computer that is connected directly to the internet. Downloading malicious programs.

\subsection{Graph models}
For some of the spreading methods, the graph of the network matters. It is important to have the right topology for the right method. Email worms need a topology that represent a social network, BGP routing worms need a topology on network level. 
--iets zeggen dat het ook belangrijk is voor de defender om zijn netwerk zo aan te passen dat het virus moeilijker kan verspreiden. Network segregation mss aanhalen ?--

\begin{description}
\item Power law
\item Small world topology
\item random graph
\item ..
\end{description}

\subsection{Models for worm propagation}
To use the models for worm propagation in our model of the FlipIt game it is suffi"ent to know the propagation speed of the worm and define after how many time units the network is defected. This will be equal to the \textit{d} parameter in our model.  
Stackelberg game: first move is from attacker, defender is follower.

@@

%What do we need for an algorithm
%\begin{description}
%\item Graph network $G = < V, E>$
%\item Graph matrix $[A]$ which is $|V| \times |V| $
%\item Attack vector $[X]$ which is $1 \times |V|$
%\item cummulative matrix $[M]$ which is $|V| \times |V|$
%\item state matrix $[T]$  which is $|V| \times |V|$
%\item Reset vector $[R]$
%\item duration \textit{d}
%\item time \textit{n}
%\item rate $\delta _{0}$ of defender and $\delta _{1}$ of attacker
%\end{description}
%
%
%
%Initialisation algorithm:
%
%
%\begin{verbatim}
%initialisatie
%	d=0
%	A=basismatrix
%	M=A^{0}
%	n=0
%	\delta_{0}
%	\delta_{1}
%	X
%	R
%	controller = defender
%	
%	
%
%	Algorithm
%	n:= n + 1;
%	Check who is in control? ( through modulo )
%	if ( defender & controller=defender)
%				d:= d + 1;
%	
%	if ( defender & controller=attacker )
%				G = X \times R  (flippen ten voordele van defender)
%				d = 0
%				controller = defender
%				
%	if ( attacker & controller=defender )
%				controller=attacker
%				..
%				
%	if ( attacker & contoller=attacker )
%				d:= d + 1
%				M = M x A
%				T = T + M
%				G = X x T
%				
%		
%\end{verbatim}

### 2015-06-21 doc/template/chap-9.tex
hoofdstuk worm propagatie aangevuld

The action of the defender consist of defending his network. The network is composed of multiple nodes. The defender can choose to flip one node, a subset of nodes or all nodes of the network. In this paper, the defender flips all the nodes of the network in every move he plays. The action of the defender can be extended to the action where the defender only flips a subset of the nodes or a single node of the network. This extension has already been investigated in the context of placing anti-virus systems in a graph. This problem is known to be NP-hard. [\todo{referentie}]. \\

@@

Even when the APT is lauched to steal sensitive information the defender can set-up a honey pot, making the attacker believe that he is stealing sensitive information, but the information in the honeypot is all fake. Honeypots emulate services or creates multiple instances of real operating systems and can pretend to have sensitive information.  Honeypots do not detect all malicious attacks so the attacker can stay unnoticed and can still provide information as feedback. This is a way how the defender can also be stealthy by providing  false information through the honeypot.\\

>>>>

The action of the defender consist of defending his network. The network is composed of multiple nodes. The defender can choose to flip one node, a subset of nodes or all nodes of the network. In this paper, the defender flips all the nodes of the network in every move he plays. The action of the defender can be extended to the action where the defender only flips a subset of the nodes or a single node of the network. This extension has already been investigated in the context of placing anti-virus systems in a graph. This problem is known to be NP-hard. \cite{NPhard}. \\

@@

Even when the APT is launched to steal sensitive information the defender can set-up a honey pot, making the attacker believe that he is stealing sensitive information, but the information in the honeypot is all fake. Honeypots emulate services or creates multiple instances of real operating systems and can pretend to have sensitive information.  Honeypots do not detect all malicious attacks so the attacker can stay unnoticed and can still provide information as feedback. This is a way how the defender can also be stealthy by providing  false information through the honeypot.\\
